---
title: "Diversity Indices 16S & ITS"
author: "Isabel Siles Asaff"
date: "2023-08-14"
output: html_document
---

## Introduction

Once data has been pre-processed and filtered alfa and beta diversity need to be calculated. Specifically in this project bacterial and fungal diversity are in two separated objects and as they were sequenced as 16s and ITS regions respectively. However, both datasets will be analyzed using the same work pipeline.

## Libraries

```{r include=FALSE}
library(phyloseq); packageVersion(phyloseq)
library(ggplot2); packageVersion(ggplot2)
library(vegan)
library(breakaway) # chao-bunge
library(microbiome) # species evenness
library(RColorBrewer)
library(tidyverse)
library(dplyr)
library(car) # ANOVA
```

## Data

```{r}
### Upload

load("ITS_filt.RData")
load("Psoil_filt.Rdata")

### Rename samples

Psoil_filt@sam_data$Treatment[Psoil_filt@sam_data$Treatment == "Control"] <- "Pi"
Psoil_filt@sam_data$Plant_Type[Psoil_filt@sam_data$Plant_Type == "No plant"] <- "Bulk Soil"
Psoil_filt@sam_data$Plant_Type[Psoil_filt@sam_data$Plant_Type == "ptxD-36"] <- "ptxDNt-36"

```

## Alpha Diversity

### Calculating Alpha Diversity Estimators 

```{r}
### Separate metadata for table

samplemeta <- as.data.frame(ITS_filtered@sam_data)

### Calculate alpha diversity using phyloseq package

# 16S
alpha_16S <- estimate_richness(Psoil_filt, 
                              split = TRUE, 
                              measures = NULL)

# ITS
alpha_ITS <- estimate_richness(ITS_filtered, 
                              split = TRUE, 
                              measures = NULL)

### Join alpha indices to metadata

alpha_16S <- cbind(samplemeta,alpha_16S)
alpha_ITS <- cbind(samplemeta,alpha_ITS)

### Calculate species evenness

Even_16S <- evenness(Psoil_filt, index = 'all',
                     zeroes = TRUE, detection = 0) 
Even_ITS <- evenness(ITS_filtered, index = 'all',
                     zeroes = TRUE, detection = 0) 

## Add evenness to existing tables  
alpha_16S <- cbind(alpha_16S,Even_16S)
alpha_ITS <- cbind(alpha_ITS,Even_ITS)

### Save files 

write.csv(alpha_16S, "~/Grad_School/Maestria/Processed_data/Alphadiversity.csv")
write.csv(alpha_ITS, "~/Grad_School/Maestria/Processed_data/Alphadiv_ITS.csv")

```

### Plotting Alpha Diversity

```{r}
### Arrange variable names in order

# 16S
Psoil_filt@sam_data$Plant_Type <- factor(Psoil_filt@sam_data$Plant_Type, levels = c("ptxDNt-36", "Wild Type", "Bulk Soil"))

Psoil_filt@sam_data$Treatment <- factor(Psoil_filt@sam_data$Treatment, levels = c("Low P", "Pi", "Phi", "Pi/Phi mix"))

# ITS
ITS_filtered@sam_data$Plant_Type <- factor(ITS_filtered@sam_data$Plant_Type, levels = c("ptxDNt-36", "Wild Type", "Bulk Soil"))

ITS_filtered@sam_data$Treatment <- factor(ITS_filtered@sam_data$Treatment, levels = c("Low P", "Pi", "Phi", "Pi/Phi mix"))

### Plot richness using microbiome

# 16S
PR_16S <- plot_richness(ITS_filtered, x="Treatment",
              measures=c("Observed","Shannon","Simpson","Chao1"),
              color="Plant_Type")+
          scale_fill_manual(breaks=c("ptxDNt-36","Wild Type","Bulk Soil"),
          labels=c("ptxDNt-36","Wild Type","Bulk soil"))+
          scale_color_manual(values = c("steelblue","lightpink","burlywood3"), 
          name = "Plant Type", 
          labels= c("ptxDNt-36","Wild Type","Bulk soil")) +
  theme_bw()

PR_16S$layers <- PR_16S$layers[-1]
PR_16S <- PR_16S + geom_point(size=4, alpha=0.7)
PR_16S

# ITS
PR_ITS <- plot_richness(ITS_filtered, x="Treatment",
              measures=c("Observed","Shannon","Simpson","Chao1"),
              color="Plant_Type")+
          scale_fill_manual(breaks=c("ptxDNt-36","Wild Type","Bulk Soil"),
          labels=c("ptxDNt-36","Wild Type","Bulk soil"))+
          scale_color_manual(values = c("steelblue","lightpink","burlywood3"), 
          name = "Plant Type", 
          labels= c("ptxDNt-36","Wild Type","Bulk soil")) +
  theme_bw()

PR_ITS$layers <- PR_ITS$layers[-1]
PR_ITS <- PR_ITS + geom_point(size=4, alpha=0.7)
PR_ITS

### Pielou: example of evenness index

# 16S
ggplot(alpha_16S, aes(x=Treatment, y=pielou, color = Plant_Type))+
  geom_point(size = 4, alpha=0.7) +
  scale_color_manual(values = c("burlywood3","steelblue","lightpink"))+
  theme_bw()

# ITS
ggplot(alpha_ITS, aes(x=Treatment, y=pielou, color = Plant_Type))+
  geom_point(size = 4, alpha=0.7) +
  scale_color_manual(values = c("burlywood3","steelblue","lightpink"))+
  theme_bw()

```

### Statistical analysis

```{r}
### Upload data 

# Only if starting in this part of the document

# 16S
alpha_16S <- read.csv("~/Grad_School/Maestria/Processed_data/Alphadiversity.csv")
row.names(alpha_16S) <- alpha_16S[,1]
alpha_16S <- alpha_16S[,-1]

# ITS
alpha_ITS <- read.csv("~/Grad_School/Maestria/Processed_data/Alphadiv_ITS.csv")
row.names(alpha_ITS) <- alpha_ITS[,1]
alpha_ITS <- alpha_ITS[,-1]

### Check data normalization

# histograms
for (i in 5:18) { 
  hist(alpha_16S[,i],
       main = i)
}
# shapiro test
for(i in 5:18){
  shapiro <- shapiro.test(alpha_16S[,i])
  normal <- ifelse(shapiro[["p.value"]]>0.05, "YES","NO")
  print(c(i,normal))
}

# repeat for ITS object

### Estimators for 16S

## Observed 

# ANOVA
ad_anova <- aov(Observed ~ Plant_Type * Treatment, 
                   data = alpha_16S) # synergistic effect
Anova(ad_anova) # not significant

## Chao1 

# ANOVA
ad_anova <- aov(Chao1 ~ Plant_Type * Treatment, 
                data = alpha_16S) # synergistic effect
Anova(ad_anova) # not significant

## Shannon

# ANOVA
ad_anova <- aov(Shannon ~ Plant_Type * Treatment, 
                data = alpha_16S) # synergistic effect
Anova(ad_anova) # not significant

## Simpson

# ANOVA
ad_anova <- aov(Simpson ~ Plant_Type * Treatment, 
                data = alpha_16S) # synergistic effect
Anova(ad_anova) # not significant 

## Pielou

# ANOVA
ad_anova <- aov(pielou ~ Plant_Type * Treatment, 
                data = alpha_16S) # synergistic effect
Anova(ad_anova) # not significant

### Estimators for ITS

## Observed

# ANOVA

ad_anova <- aov(Observed ~ Plant_Type * Treatment, 
                   data = alpha_ITS) # synergistic effect
Anova(ad_anova) # signifcant in plant:treatment

# Test homogeneity of variance

plot(ad_anova, 1) # check outliers
leveneTest(Observed ~ Plant_Type * Treatment, 
           data = alpha_16S) # higher than 0.05 so it meets ANOVA assumption

# Tukey HSD 

ad_Tukey <- TukeyHSD(ad_anova) # bulk soil Pi/Phi is statistically different from several samples

## Chao1

# ANOVA
ad_anova <- aov(Chao1 ~ Plant_Type * Treatment, 
                data = alpha_ITS) # synergistic effect
Anova(ad_anova) # significant in plant:treatment

# Test homogeneity of variance

plot(ad_anova, 1) # check outliers
leveneTest(Observed ~ Plant_Type * Treatment, 
           data = alpha_ITS) # higher than 0.05 so it meets ANOVA assumption

# Tukey HSD 

ad_Tukey <- TukeyHSD(ad_anova) # bulk soil Pi/Phi is statistically different from several samples

## Shannon

# ANOVA
ad_anova <- aov(Shannon ~ Plant_Type * Treatment, 
                data = alpha_ITS) # synergistic effect
Anova(ad_anova) # not significant

## Simpson

# ANOVA
ad_anova <- aov(Simpson ~ Plant_Type * Treatment, 
                data = alpha_ITS) # synergistic effect
Anova(ad_anova) # not significant 

## Pielou

# ANOVA
ad_anova <- aov(pielou ~ Plant_Type * Treatment, 
                data = alpha_ITS) # synergistic effect
Anova(ad_anova) # not significant

```

## Beta Diversity

### Distance and ordination calculations

```{r}
### 16S

## Distances

# Bray Distance
bray_16S <- phyloseq::distance(Psoil_filt, 
                                method = "bray", type = "sample")

# Jaccard: Presence/Absence 
jacc_16s <- phyloseq::distance(Psoil_filt, 
                                method = "jaccard", binary = TRUE,
                                type = "sample") # binary argument needed 

## Ordination methods

# NMDS
B_ord16S <- ordinate(ITS_filtered,"NMDS", distance = bray_16S)

# PCoA
J_ord16S <- ordinate(ITS_filtered,"PCoA", distance = jacc_16s)

# Distance and ordination methods can be iterated


### ITS

## Distances

# Bray Distance
bray_bdiv <- phyloseq::distance(ITS_filtered, 
                                method = "bray", type = "sample")

# Jaccard: Presence/Absence 
jacc_bdiv <- phyloseq::distance(ITS_filtered, 
                                method = "jaccard", binary = TRUE,
                                type = "sample") # binary argument needed 

## Ordination methods

# NMDS
bray_ord <- ordinate(ITS_filtered,"NMDS", distance = bray_bdiv)

# PCoA
jacc_ord <- ordinate(ITS_filtered,"PCoA", distance = jacc_bdiv)

```

### Plotting distances 

```{r}

### 16S

## Bray

p_bray <- plot_ordination(Psoil_filt, B_ord16S, "samples", 
                          color = "Plant_Type",
                          shape = "Treatment")+
  scale_color_manual(values = c("steelblue","lightpink","burlywood3"), name = "Plant Type")+
  theme_bw()+
  geom_point(size=3)

shape_names <- c("Pi","Low P","Phi", "Pi/Phi mix")
p.shape <- 15:(15 + length(shape_names) - 1)
names(p.shape) <- shape_names
p.shape["Treatments"] <- 4

p_bray <- p_bray +
  scale_shape_manual(values = p.shape)

p_bray <- p_bray +
  geom_polygon(aes(fill=Plant_Type),alpha = 0.2)+
  scale_fill_manual(values = c("steelblue","lightpink","burlywood3"), 
                    name = "Plant Type")

p_bray

## Jaccard

p_jacc <- plot_ordination(Psoil_filt, J_ord16S, "samples", 
                          color = "Plant_Type",
                          shape = "Treatment")+
  scale_color_manual(values = c("steelblue","lightpink","burlywood3"), name = "Plant Type")+
  theme_bw()+
  geom_point(size=3)

shape_names <- c("Pi","Low P","Phi", "Pi/Phi mix")
p.shape <- 15:(15 + length(shape_names) - 1)
names(p.shape) <- shape_names
p.shape["Treatments"] <- 4

p_jacc <- p_jacc +
  scale_shape_manual(values = p.shape)

p_jacc <- p_jacc +
  geom_polygon(aes(fill=Plant_Type),alpha = 0.2)+
  scale_fill_manual(values = c("steelblue","lightpink","burlywood3"), 
                    name = "Plant Type")

p_jacc


### ITS

## Bray

p_bray <- plot_ordination(ITS_filtered, bray_ord, "samples", 
                          color = "Plant_Type",
                          shape = "Treatment")+
  scale_color_manual(values = c("steelblue","lightpink","burlywood3"), name = "Plant Type")+
  theme_bw()+
  geom_point(size=3)

shape_names <- c("Pi","Low P","Phi", "Pi/Phi mix")
p.shape <- 15:(15 + length(shape_names) - 1)
names(p.shape) <- shape_names
p.shape["Treatments"] <- 4

p_bray <- p_bray +
  scale_shape_manual(values = p.shape)

p_bray <- p_bray +
  geom_polygon(aes(fill=Plant_Type),alpha = 0.2)+
  scale_fill_manual(values = c("steelblue","lightpink","burlywood3"), 
                    name = "Plant Type")

p_bray

# Jaccard
p_jacc <- plot_ordination(ITS_filtered, jacc_ord, "samples", 
                          color = "Plant_Type",
                          shape = "Treatment")+
  scale_color_manual(values = c("steelblue","lightpink","burlywood3"), name = "Plant Type")+
  theme_bw()+
  geom_point(size=3)

shape_names <- c("Pi","Low P","Phi", "Pi/Phi mix")
p.shape <- 15:(15 + length(shape_names) - 1)
names(p.shape) <- shape_names
p.shape["Treatments"] <- 4

p_jacc <- p_jacc +
  scale_shape_manual(values = p.shape)

p_jacc <- p_jacc +
  geom_polygon(aes(fill=Plant_Type),alpha = 0.2)+
  scale_fill_manual(values = c("steelblue","lightpink","burlywood3"), 
                    name = "Plant Type")

p_jacc
```

### Statistical analysis: PERMANOVA

```{r}
### Extract data frames from phyloseq object

## count table: ASVs

ASVs_ITS <- as.data.frame(ITS_filtered@otu_table)
ASVs_16S <- as.data.frame(Psoil_filt@otu_table)

## metadata or Environment

SoilMeta <- read.csv("~/Grad_School/Maestria/Processed_data/metadata.csv")

### Basic Adonis

## 16S

# Bray distance
adonis2(ASVs_16S ~ Treatment * Plant_Type, 
        data=SoilMeta, permutations = 999,
        method = "bray") # no significant differences

# Cao distance
adonis2(ASVs_16S ~ Treatment * Plant_Type, 
        data=SoilMeta, permutations = 999,
        method = "cao")
# plant type marginal difference p = 0.096

# Jaccard
adonis2(ASVs_16S ~ Treatment * Plant_Type, 
        data=SoilMeta, permutations = 999,
        method = "jaccard", binary = TRUE) # add binary 
# no significant difference

## ITS

# Bray distance
adonis2(ASVs_ITS ~ Treatment * Plant_Type, 
        data=SoilMeta, permutations = 999,
        method = "bray") # no significant differences

# Cao distance
adonis2(ASVs_ITS ~ Treatment * Plant_Type, 
        data=SoilMeta, permutations = 999,
        method = "cao")
# plant type marginal difference p = 0.096

# Jaccard
adonis2(ASVs_ITS ~ Treatment * Plant_Type, 
        data=SoilMeta, permutations = 999,
        method = "jaccard", binary = TRUE) # add binary 
# no significant difference

```

### Calculating Species Scores

```{r}
### 16S ###

NMDS_16S <- metaMDS(ASVs_16S, distance = "bray")
Scores_16S <- scores(NMDS_16S)
Sites_16S <- as.data.frame(Scores_16S[["sites"]])

## Plot scores 

ggplot(data=Sites_16S) + 
  stat_ellipse(aes(x=NMDS1,y=NMDS2,colour=SoilMeta$Plant_Type), type = "norm", level = 0.50) +
  geom_point(aes(x=NMDS1,y=NMDS2,shape=SoilMeta$Treatment,colour=SoilMeta$Plant_Type),size=3) + 
  theme_bw() 

### ITS ###

NMDS_ITS <- metaMDS(ASVs_ITS, distance = "bray")
Scores_ITS <- scores(NMDS_ITS)
Sites_ITS <- as.data.frame(Scores_ITS[["sites"]])

## Plot scores 

ggplot(data=Sites_ITS) + 
  stat_ellipse(aes(x=NMDS1,y=NMDS2,colour=SoilMeta$Plant_Type), type = "norm", level = 0.50) +
  geom_point(aes(x=NMDS1,y=NMDS2,shape=SoilMeta$Treatment,colour=SoilMeta$Plant_Type),size=3) + 
  theme_bw() 
```


